\thetitle

### 2013-04-17 tfpie/tfpie.tex
Start on section 3.

The semester starts with the minimal conceptual change from the
previous semester: the addition of objects (same IDE, same syntax,
same semantics [just a new kind of value], same interactive video
games, etc.).

>>>>

On the first day of the second semester, we introduce a single
linguistic concept to an otherwise unchanged context of the previous
semester.

An object is a new kind of value that can, as a first cut, be
understood as a pairing together of two familiar concepts: data and
functionality.

\begin{itemize}
\item An object is like a structure in that it has a fixed number of
fields, thus an object (again, like a structure) can represent
compound data. But unlike a structure, an object contains not just
data, but functionality too;

\item An object is like a (set of) function(s) in that it has behavior—it computes; it is
not just inert data.
\end{itemize}

This suggests that objects are a natural ﬁt for well-designed programs
since good programs are organized around data definitions and
functions that operate over such data. An object, in essence, packages
these two things together into a single programming apparatus. This
has two important consequences:

\begin{enumerate}

\item Students already know how to design programs oriented around objects.

Since objects are just the combination of two familiar concepts that
students already use to design programs, they already know how to
design programs around objects, even if they have never heard the term
``object'' before. Our message is the better you are at programming
with functions, the better you will be at programming with objects,
and \emph{vice versa}.

\item Objects enable new kinds of abstraction and composition.

Although the combination of data and function may seem simple, objects
enable new forms of abstraction and composition. That is, objects open
up new approaches to the construction of computations. By studying
these new approaches, we can distill new design principles. Because we
understand objects are just the combination of data and function, we
can understand how all of these principles apply in the familiar
context of programming with functions. In short, the better you are at
programming with objects, the better you will be at programming with
functions.

\end{enumerate}

\subsection{The basics of objects}

To begin with, we introduce the notion of a \emph{class definition},
which can be thought of at first as a structure definition in that
it defines a new class of compound data.  A class is defined using the
{\tt define-class} form:
\begin{verbatim}
    (define-class posn% (fields x y))
\end{verbatim}
An \emph{object} is a value that is a member of this class of data,
which can be constructed with the {\tt new} keyword, a class name, and
the appropriate number of arguments for the fields of the object:
\begin{verbatim}
    (new posn% 3 4)
\end{verbatim}
An object understands some set of \emph{messages}.  Simple
structure-like objects understand messages for accessing their fields
and message are sent by using the {\tt send} keyword, followed by an object,
a message name, and some number of arguments:
\begin{verbatim}
    (send (new posn% 3 4) x) ;=> 3
    (send (new posn% 3 4) y) ;=> 4
\end{verbatim}
The {\tt send} notation is simple, but syntactically heavy.  Once
students are comfortable with the {\tt send} form, we introduce a
shorthand to make it more convenient by writing {\tt (x . m)} for {\tt
  (send x m)}.  The \emph{dot notation} can be nested, so {\tt (x . m
  . n)} is shorthand for {\tt (send (send x m) n)}.  (The approach of
introducing a simple, uniform syntax and later introducing a
convenient shorthand that would have been confusing to start with
follows the approach of introducing {\tt cons} and later {\tt list}
and {\tt quote} in the first semester.)

It is possible to endow objects with functionality by defining \emph{methods},
which extend the set of messages an object understands.  A method definition
follows the same syntax as a function definition, but is located inside of a
class definition.  Here is a more complete development of the {\tt posn\%} class
that includes a couple of methods:
\begin{verbatim}
    ;; A Posn is a (new posn% Number Number),
    ;; which represents a point on the Cartesian plane
    (define-class posn% (fields x y)

      ;; Posn -> Number
      ;; Distance between this posn and that posn
      (check-expect ((new posn% 0 0) . dist (new posn% 3 4)) 5)
      (define (dist that)
        (sqrt (+ (sqr (- (this . x) (that . x)))
                 (sqr (- (this . y) (that . y))))))

      ;; -> Number
      ;; Distance of this posn from the origin
      (check-expect ((new posn% 0 0) . dist-origin) 0)
      (check-expect ((new posn% 3 4) . dist-origin) 5)
      (define (dist-origin)
        (this . dist (new posn% 0 0))))
\end{verbatim}

This class definition defines a new class of values which are {\tt
  posn\%} objects.  Such objects are comprised of two numeric values
and understand the messages {\tt x}, {\tt y}, {\tt dist}, and {\tt
  dist-origin}.  Unit tests have been included with each method
definition, following the principles of the design recipe studied in
the first semester.  In fact, the {\tt check-expect} mechanism works
exactly as it did before.

Methods can be defined to consume any number of arguments, but they
are implicitly parameterized over {\tt this}, the object that was sent
the message.

\subsection{Dispatch}

Unions, and recursive unions in particular, are a fundamental kind of
data definition that students are well-versed in from the previous
semester.  A fundamental early lesson is how to represent (recursive)
unions using classes and how to write recursive methods.  As an
example, here is a simple class definition for lists of numbers
(an archetypical recursive union data definition) that includes
a method for computing the length of a list:
\begin{verbatim}
    ;; A list of numbers (LoN) is one of:
    ;; - (new empty%)
    ;; - (new cons% Number LoN)

    (define-class empty%
      ;; Length of this empty list
      (check-expect ((new empty%) . len) 0)
      (define (len) 0))

    (define-class cons% (fields first rest)
      ;; Length of this cons list
      (check-expect ((new cons% 7 (new empty%)) . len) 1)
      (define (len)
        (add1 (this . rest . len))))
\end{verbatim}

The structure of this data definition is analagous to the approach of
the previous semester but this example brings to light an important
difference with the functional approach.

%% The semester starts with the minimal conceptual change from the
%% previous semester: the addition of objects (same IDE, same syntax,
%% same semantics [just a new kind of value], same interactive video
%% games, etc.).
